The dynamics of risk perception in a Mediterranean agroecosystem

### 2020-03-06 manuscript/manuscript.tex
typos

\title{The dynamics of risk perception in a Mediterranean agroecosystem \thanks{Grants or other notes about the article that should go on the front page
should be placed here. General acknowledgments should be placed at the
end of the article.} }

    \titlerunning{Risk perception in a Mediterranean agroecosystem}

\author{  Nicolas Gauthier 1 \and  }

@@

\institute{
        Nicolas Gauthier 1 \at
     Laboratory of Tree-Ring Research \& School of Geography and Development,
 University of Arizona \\
     \email{\href{mailto:ngauthier@email.arizona.edu}{\nolinkurl{ngauthier@email.arizona.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
Small-scale agriculturalists in the Mediterranean Basin rely on multiple
strategies including diversification, intensification, and storage to
maintain a stable food supply in the face of environmental uncertainty.
Each of these strategies requires farmers to make specific resource
allocation decisions in response to environmental risks and is thus
sensitive to variability in both the spatiotemporal pattern of risk and
the ability of farmers to perceive that pattern. In this chapter, I
present an agent-based model of a Mediterranean agroecosystem. By
driving the model with realistic environmental dynamics derived from
simulations of mid-Holocene Mediterranean climate, and by allowing the
psychology of risk perception to vary among individual farmers, I
explore the hidden vulnerabilities of traditional risk-management
strategies to periods of rapid climate change. I show that even when
farmers are able to manage risk ``optimally'' in light of past
experience, unanticipated changes in the spatiotemporal pattern of
rainfall can still lead to major food shortfalls.
\\
\keywords{
        key \and
        dictionary \and
        word \and
    }

    \subclass{
                    MSC code 1 \and
                    MSC code 2 \and
            }

@@

Over the past 10,000 years, subsistence farmers in the Mediterranean
basin have developed sophisticated strategies to maintain stable food
supplies given uncertain rainfall. Their suite of strategies include
practices like crop diversification, storage, and exchange (Halstead and
O'Shea 1989)). These risk-management strategies all represent forms of
\emph{regulatory feedback} used to manage complex socio-ecological
systems. Regulatory feedbacks convert a system's high-variance inputs
into low-variance outputs by dynamically monitoring the outputs and
adjusting internal properties of the system ({\textbf{???}}).
Risk-managing strategies require farmers to make specific resource
allocation decisions in response to environmental risks, and as
regulatory feedbacks they are thus sensitive to variability in both the
spatiotemporal patterns of risk and the ability of farmers to perceive
and act on those patterns.

Crop diversification is an excellent example of a widespread and
effective regulatory feedback strategy (Figure 1). Relying on a mix of
food types with different climatic tolerances is an efficient way to
maintain a robust food supply ({\textbf{???}},Anderies2006). In the
Mediterranean, land-use strategies involving a diversified portfolio of
wheat and barley have been employed by even the earliest sedentary
farmers, and continue to be used to this day (Gould
1963,Slafer1999,Abbo2009a,Marston2011190). Wheat generally has higher
yields but is sensitive to water deficits, while barley has lower yields
and is drought tolerant. Planting a mix of wheat and barley, either in
the same plot or in a combination of plots, is an effective means of
diversifying grain supplies. By adjusting the ratio of wheat to barley,
farmers can adapt to changing drought risks.

The secret to the success of earliest farmers on the mediterrenan
littoral was in their manipulation of the local ecosystem to manage
environmental risks. Early farmers constructed a complex agroecosystem
to manage risk. They needed to maintain stable crop yields in face of
uncertain drought risks. Yield stability is very important for
traditional agriculturlal commmunities not connected to larger bulk food
networks (Abbo et al 2010) neolithic crop package -- weiss zohary 2011
cite abbo et al 2010? for more stuff about ladraces Wheat/barley systems
example of system that combines low yield, low wirks crop with high
yield high risk crop, an effective context paut et al 2019 These
strategies work when risks are knonw and constant, but may be more
susceptible to periods of changing climateic risks so understanding how
people of the past adapted to change is important to us today

Are these crop-diversification strategies vulnerable to the same
dynamics as other social-ecological systems with similar regulatory
feedbacks mechanisms? One way to address this is to model the influence
of imperfect monitoring and biased decision making in uncertain
environments. In this study, I accomplish this by answering two main
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  What was the temporal pattern of climate and climate change during the
  Holocene in the eastern Mediterranean?
\item
  How well could crop diversification strategies cope with this pattern,
  given farmers' imperfect risk perception and decision-making?
\end{enumerate}

Questions: How would the earliest farmers have made decisions (dealt
with uncertainty?) about crop diversification? ie what were the
different algorithms? -- better understand the decision context What are
the consequences of theses different decisions (i.e.~how well did they
perform) at different points in the history of the mediterranean How
were farmers able to detect changes in the environment, rather than
unconsciously adapt to them? what are the different consequences we'd
expect for each THIS is the most important question because with it I
can frame the results around and individual's perception of droughts in
the neolithic, and the populaiton-level distribution of drought
perceptions at given times

I couple a climate model with high temporal resolution to a simple game
theoretic model of agricultural decision making under uncertainty, in
order to examine the performance of optimal and suboptimal wheat-barley
diversification practices. I model the year-to-year crop diversification
strategies made by farmers as an iterated game of fictitious play
against nature (Gould 1963). Farmers are boundedly rational, in that
they seek to maximize their objective functions given the perceived
probabilities of different actions by nature, but work with flawed
mental models of climate-related risks due to imperfect recall of past
events. Furthermore, by using a climate model to represent the actions
of nature, I can more precisely capture the characteristic
autocorrelated patterns of rainfall variability, rather than simply
drawing rainfall values from a static distribution. This framework
allows me to address the sensitivity of diversification strategies to
changes and climatic variability and imperfect human perceptions of
those changes.

@@

The basic decision-making problem facing a farmer seeking to diversify
their crops can be thought of as a game against nature. We can represent
this simplified decision context as a ``game'' in a game theorietic
context. cite luce raiffa 1989 for game against nature, and milnor 1952,
Agrawal1968, cassidy1971, gould 1963 The ``game'' in this context is the
farmer's decision of which crops to plant and in what proportions, given
uncertianty in the future ``state-of-nature'', and nature varies between
several possible states such as dry and wet years. payoff matrix goes
here in reality continuous, but cite behavioral econ for people often
intuitively solve an easier problem when faced with a complex real-world
situation The exact values in the payoff matrix here are less important
than the relative payoffs in each quadrant. We can think of this game as
being the culturally-inherited object. our question is how do
individuals make decisions here? so in this case the cultural norm would
be to mix wheat and barley, but the learning is in the exact ratio --
both individual and social learning influence perceptions of eindividual
vs social learning tucker 2007

Estimates of yield volume (t/ha) for each crop type in wet and dry years
were derived from isotopic studies of ancient wheat and barley samples
from archaeological sites in the Mediterranean (Araus, Slafer, and
Romagosa 1999) (Table 1).

\begin{table}
\centering
\caption{Estimated yields for ancient wheat and barley varieties derived from [@Slafer1999].}
\begin{tabular}{|l|l|l|}
\hline

 & Dry Year & Normal Year \\ \hline
Barley Yield & 0.93 & 1.18 \\ \hline
Wheat Yield & 0 & 1.60 \\ \hline

\end{tabular}
\end{table}

How should an early farmer make a decision in light ot this uncertainty?
The basic logic of crop diversitication is Portfolio of crops ala modern
portfolio theory: Blank2001 Sometimes best to think of nature as a
sentient opponent out to get you, and play strategically based on that
assumption. Gould 1963, Beckenkamp 2008 quick overview of criteria
without risk, then say that a better option is to estimate the
probabilities of the different state of nature Agrawal1968 talks about
the strategies in the context of agriculture being a game against nature
here I can slip in the cool thing about acting like the weather is out
to get you being a good idea in states of complete uncertainty
transition -- the problem becomes easier if you have at least someI idea
of nature's moves , because then you cna work to get the highest yeilds
given the risk of drought Decision strategies as choosing different
points -- plot of decisions

\begin{align}
a^2+b^2=c^2
\end{align}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Expected wheat and barley yields under increasing drought risk
with the point of indifference highlighted.}
\end{figure}

If an individual can learn about their environment, such as the risk of
drought in any given year, then they can behave more rationally by
trying to maximize their subjective expected utility If the probability
distributin of nature's moves is known, the farmer can choose the crop
mix that simply maximizes the expected crop yields given the empirical
frequency distirubtion of nature's moves. In the language of game
theory, this strategy is knoan as a ``game of ficticious play'' against
nature. This is an effective risk-managing strategy, but like all
similar strategies that adapt to a specific pattern of variability, it
is vulnerable to changes in the pattern of variability (Janssen eta al
2007) This strategy works well when the environment is stationary --
nature plays from a fixed probability distirubtion, but is vulnerable to
environmental nonstationarity. That is, when the mean or higher order
moments of the rainfall distribution shift, playing a game of fictitious
play can backfire because you're too stuck in your ways while the world
changes around you. Thus it is important to understand the dynamics of
risk perception -- how subjective risks, and thus subjective expected
utility -- rise and fall in uncertain environments.

\hypertarget{sec:2}{%
\section{The dynamics of risk perception}\label{sec:2}}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-2.pdf}

\hypertarget{drought-risks-in-the-middle-holocene}{%
\section{Drought risks in the middle
Holocene}\label{drought-risks-in-the-middle-holocene}}

\hypertarget{climate-risks-and-risk-perception}{%
\section{Climate Risks and Risk
Perception}\label{climate-risks-and-risk-perception}}

Using the bias-corrected climate model output, I divided each model year
into dry years and wet years. A dry year was any year where less than
300mm of rain fell during the wet season (October-March), the threshold
below which wheat crops will generally fail ({\textbf{???}}), and a
wet/normal year was defined as any year above this threshold.

Given the modeled patterns of wet and dry years, drought risk for any
particular year was defined as the proportion of the previous 50 years
that were dry years

\begin{equation}
    P_{dry} = \frac{\sum_{n=t-1}^{t-50} precip_n < 300}{50},
\end{equation} where \(t\) is the current time step. The 50 year time
span was selected to approximate the accumulated observational knowledge
of an individual farmer and their immediate household. As a result, this
approach does not allow for accumulated social learning, although it
could easily be extended to do so in future studies.

\hypertarget{climate-modeling-and-bias-correction}{%
\subsection{Climate Modeling and Bias
Correction}\label{climate-modeling-and-bias-correction}}

Estimates of drought risk in the eastern Mediterranean during the the
past 10,000 years were derived from outputs of the TraCE-21k
paleoclimate simulation available on the National Center for Atmospheric
Research's Earth System Grid repository
(\url{https://www.earthsystemgrid.org}). TraCE-21K is a state-of-the-art
simulation that uses a coupled atmosphere-ocean general circulation
model (GCM) to recreate the transient response of the global climate
system to orbital parameters and greenhouse gasses over the past 22,000
years from the Last Glacial Maximum to the present (Figure 2) (He 2011).
It generates physically consistent spatiotemporal climate dynamics,
driven by current best estimates of climate forcings (e.g.~orbit,
greenhouse gasses, glacial meltwater flux). The model simulates these
dynamics on a six hourly timescale, and model outputs are archived at a
monthly resolution.

TraCE-21k simulation outputs for the past 10,000 years were
bias-corrected using the CDF-t method and observed precipitation from a
weather station in near the town of Salihli in western Turkey. This
location was selected due to its long (\textasciitilde{}70yr)
observational record, the representativeness of western Anatolian
climate for the greater eastern Mediterranean, and its proximity to
archaeological sites with comparative evidence of crop diversification
strategies.

Now incorporate realistic drought variability from simulations.

\begin{verbatim}
## Loading required namespace: ncdf4
\end{verbatim}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Annual risk of wheat crop failure due to drought, averaged by
fifty-year period}
\end{figure}

\hypertarget{the-model}{%
\section{The Model}\label{the-model}}

Run an experiment. Simulate a time series with varying drought
frequency. This is a stepwise nonstationary process

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-12-1.pdf}

Redo these plots so its a single person, one who was born in a drought
period then experiences an ameliorarion and another who experiences a
shift to drought conditions later in life? basically this section neds
to work more with individuals and explore that stuff

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-14-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-15-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-16-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-18-1.pdf}

These two plots begin to look more alike the more you sample, which
makes sense because its an ergodic system.

Note in these plots above and below we're summing all the wheat and
barley in the population, compared to below which shows the wheat
proportions of individuals
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-19-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-19-2.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-20-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-20-2.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-21-1.pdf}
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-21-2.pdf}

\begin{verbatim}
## Picking joint bandwidth of 0.0146
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-22-1.pdf}

\begin{verbatim}
## Picking joint bandwidth of 0.0204
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-22-2.pdf}

make this a function to calculate payoffs for all wheat or all barley
strategies as a comparison
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-24-1.pdf}

\section{Results and Discussion}

The bias-corrected wet season precipitation outputs from TraCE-21k show
a relatively stationary distribution over the past 10,000 years, with
slightly drier conditions during the early-mid Holocene transition at
c.a. 8ka BP (Figure 5). The utility of a diversified wheat-barley crop
is immediately apparent; rainfall often fails to meet the 300mm
threshold for a productive wheat crop during the entire 10,000 year
span, but drops below the threshold for barley only a handful of years.

The risk of crop failure due to drought in any given year varies between
10\% and 20\%, punctuated by one to four century-long events where risk
drops as low as 6\% and or exceeds 25\% (Figure 6). Centuries of high
drought risk cluster around known periods of regional and global climate
deterioration, such as at 8.6ka BP and 4.2ka BP.

Given the average drought risks estimated from TraCE-21k and assuming
perfect monitoring and recall of recent droughts (i.e. \(\lambda = 0\)),
farmers intent on minimizing the risk of crop failure would be expected
to plant about 65\% barley to 35\% wheat on average. This proportion is
within the range estimated from macrobotanical remains at a nearby
archaeological site ({\textbf{???}}). Yield maximizing farmers instead
play a pure strategy profile, planting wheat if the perceived drought
risk is less than 0.68 and barley if it is more. Because the
GCM-simulated drought risk never rises above this threshold, a
yield-maximizing farmer will always plant a wheat monocrop.

Examining the long-term dynamics of crop yields among all
decision-making strategies and psychological profiles reveals
variability in yields due to different decision making preferences is
far greater variability due to differences in risk perception (Figure
7). Risk-minimizing strategies sacrifice productivity for
predictability, consistently lowering the mean and variance of crop
yields over time. Risk-minimizing strategies are more sensitive to risk
perception than yield-maximizing strategies; varying the memory decay
rate parameter \(\lambda\) had no effect on the optimal crop allocation
for yield maximizers.

During periods of climatic stability, allowing past experiences to
influence decision making helps farmers minimize the impacts of
\emph{predictable} drought. But past experiences are less informative
during periods of rapid climate change, and even farmers who manage risk
``optimally'' experience major food shortfalls. Climatically-induced
variability in food supplies consistently surpasses that from
differences in risk-perception psychology, but not that from different
risk aversion preferences. This finding suggests that the efficacy of
risk managing strategies that rely on regulatory feedbacks is limited to
periods of relative climatic stability. This is consistent with the
observation from the Roman Period Mediterranean that while minor food
crises were common in the ancient world, extreme famines were rare
(Garnsey 1989).

Statistically downscaling palaeoclimate simulations and coupling them to
game theoretic models of decision making under uncertainty is a simple
way to better understand the patterning of climate changes and the
regulatory feedbacks farmers use to adapt to them. To further explore
these patterns, future work should incorporate additional
risk-minimizing strategies such as storage and exchange. Food storage
and exchange can also be thought of as forms of diversification, the
former in the time and the latter in space ({\textbf{???}}). Food
exchange would be of particular interest from a game theoretic
perspective, as incomplete information with respect to an exchange
partner's crop diversification and storage practices can lead to
situations of moral hazard.

Think about the importance of the population's age structure, becuase
that's going to impact the collective memory of drought.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Slafer1999}{}%
Araus, J. L., G. A. Slafer, and I. Romagosa. 1999. ``Durum wheat and
barley yields in antiquity estimated from 13C discrimination of
archaeological grains: a case study from the Western Mediterranean
Basin.'' \emph{Australian Journal of Plant Physiology} 26 (4): 345.
\url{https://doi.org/10.1071/PP98141}.

\leavevmode\hypertarget{ref-Garnsey1989}{}%
Garnsey, Peter. 1989. \emph{Famine and Food Supply in the Graeco-Roman
World: Responses to Risk and Crisis}. Cambridge University Press.

>>>>

\title{The dynamics of risk perception in a Mediterranean agroecosystem }

    \titlerunning{The dynamics of risk perception in a Mediterranean agroecosystem}

\author{  Nicolas Gauthier \and  }

@@

\institute{
        Nicolas Gauthier \at
     Laboratory of Tree-Ring Research \& School of Geography and Development, University of Arizona \\
     \email{\href{mailto:ngauthier@arizona.edu}{\nolinkurl{ngauthier@arizona.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\begin{abstract}
Small-scale agriculturalists in the Mediterranean Basin rely on multiple strategies including diversification, intensification, and storage to maintain a stable food supply in the face of environmental uncertainty. Each of these strategies requires farmers to make specific resource allocation decisions in response to environmental risks and is thus sensitive to variability in both the spatiotemporal pattern of risk and the ability of farmers to perceive that pattern. In this chapter, I present a simple agent-based model of a Mediterranean agroecosystem. By driving the model with realistic environmental dynamics derived from simulations of mid-Holocene Mediterranean climate, and by allowing the psychology of risk perception to vary among individual farmers, I explore the hidden vulnerabilities of traditional risk-management strategies to periods of rapid climate change. I show that even when farmers are able to manage risk ``optimally'' in light of past experience, unanticipated changes in the spatiotemporal pattern of rainfall can still lead to major food shortfalls.
\\
\keywords{
        Bayesian agents \and
        game against nature \and
        crop diversification \and
        drought risk \and
        rapid climate change \and
    }

@@

The distinctive climate and ecology of the Mediterranean basin afforded both challenges and opportunities to the earliest farming communities. Water was the primary limiting resource for these traditional agroecosystems. Agricultural droughts where growing season precipitation is low enough to cause crop failures are a constant threat. Precipitation is highly variable in space and time and droughts are difficult to predict with certainty. How were Neolithic farmers able to adapt to and even thrive in such an uncertain environment?

Over the past 10,000 years, small-scale subsistence farmers have relied on a suite of strategies to maintain stable food supplies given uncertain rainfall. These strategies include practices like crop diversification, storage, mobility, and exchange (Halstead and O'Shea 1989). Crop diversification in particular is an excellent example of a widespread and effective risk management strategy that is well suited to Mediterranean agroecosystems. In the Mediterranean, land-use strategies involving a diversified portfolio of wheat and barley have been employed by the earliest sedentary farmers and continue to be used to this day (Gould 1963; Araus, Slafer, and Romagosa 1999; Abbo, Lev-Yadun, and Gopher 2010; Weiss and Zohary 2011; Marston 2011). Relying on a mix of food types with different climatic tolerances is an efficient way to maintain a robust food supply (Helmers, Yamoah, and Varvel 2001; Anderies 2006). Wheat is high yielding but drought sensitive, while barley is lower yielding but drought tolerant. Planting a mix of high-yield, high-risk and low-yield, low-risk crops, either in the same plot or in a combination of plots, is an effective means of diversifying the annual supply of staple food crops (Paut, Sabatier, and Tchamitchian 2019). By dynamically adjusting the ratio of wheat to barley in their fields, farmers can adapt to a variety local climate conditions with different drought risks.

Risk-management strategies like crop diversification require farmers to make specific resource allocation decisions in response to specific environmental risks. As such, they are sensitive to variability in both the spatiotemporal patterns of risk and the ability of individual decision-makers to perceive and act on those patterns. Here, I focus on two main questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  How likely were droughts to occur each year in the eastern Mediterranean, and how did these risks change over the Holocene?
\item
  How would Neolithic farmers have perceived these changing risks, and what were the consequences for Neolithic farmers' collective ability to manage them?
\end{enumerate}

To address these questions, I first use results from a long-term paleoclimate simulation to estimate the changing risk of agricultural droughts in the eastern Mediterranean over a 4,000-year period in the early to middle Holocene. Then, I use a simulated population of ``belief-based'' Bayesian agents to explore how well individual farmers would have been able to perceive these long-term changes in drought risks given their finite life experiences and limited capacity to process information. This computational approach allows for a more nuanced understanding of the vulnerability of risk-management strategies to unpredictable climatic variability.

@@

The basic decision-making problem facing a farmer seeking to diversify their crops can be thought of as a ``game against nature'' (Milnor 1952; Gould 1963; Agrawal and Heady 1968; Cassidy, Field, and Kirby 1971; Luce and Raiffa 1989). In the context of decision theory, the ``game'' is the farmer's decision of which crops to plant and in what proportions, given uncertainty in the weather (or the ``state of nature'') in a given year. Assume that farmers are working with a simplified representation of reality and intuitively solve an easier problem when faced with a complex real-world situation (Simon 1990). That is, rather than a continuum of possible states of nature, farmers only care about two categories -- years when it is too dry to plant wheat and years when it is not.

\hypertarget{the-payoff-matrix}{%
\subsection{The payoff matrix}\label{the-payoff-matrix}}

This simplified decision problem can take the form of a payoff matrix in which the rows represent the moves of the farmer (wheat or barley) and the columns are the moves of nature (dry or normal, Table 1). The matrix is populated with realistic estimates of crop yields derived from isotopic analyses of grain residues at Neolithic sites (Araus, Slafer, and Romagosa 1999). However, the absolute yields are less important for the decision-making problem than the relative sums of each row and column. Given this payoff matrix, what strategies might a farmer use to ``win'' this game and maximize their yields?

\begin{table}
\centering
\caption{Estimates of yield volume (t/ha) for prehistoric wheat and barley varieties derived from (Slafer et al. 1999). The absolute values here are less important for decision-making than the relative values across each row and column.}
\begin{tabular}{|l|l|l|}
\hline
 & Dry Year & Normal Year \\ \hline
Barley Yield & 0.93 & 1.18 \\ \hline
Wheat Yield & 0 & 1.60 \\ \hline
\end{tabular}
\end{table}

There are several decision criteria a farmer might use in this situation (Milnor 1952; Agrawal and Heady 1968). A risk neutral farmer seeks only to maximize yields in a normal year by planting wheat and hoping for the best, as wheat is the highest yielding crop overall. But, this strategy risks starvation in drought years.

Instead it is often rational to assume ``nature's'' moves are decided by sentient being bent on one's ruin, and to play strategically as if the worst-case scenario will occur (Gould 1963; Beckenkamp 2008). In the particular form of the wheat-barley game presented in Table 1, a risk sensitive farmer would assume that droughts are inevitable and plant barley to guarantee a minimum acceptable harvest even in the worst-case scenario. This cautious strategy is beneficial in situations of complete uncertainty, but farmers miss out on the high yields they would have received from planting wheat if a drought ultimately does not occur.

\hypertarget{subjective-expected-yields}{%
\subsection{Subjective expected yields}\label{subjective-expected-yields}}

Playing this ``game'' over many years allows farmers to learn how probable each state of nature is to occur and to adjust their choices accordingly. This strategy is known as playing a game of ``fictitious play'' against nature. If the empirical frequency distribution of wet and dry years is known, the farmer can multiply the crop yields in Table 1 by the probability of each state of nature occurring and plant the crop with the highest \emph{expected} yield (Upton 1987). If the probability of a dry year is denoted \(\theta\), then the expected barley yield is \(\theta \times 0.93 + (1 - \theta) \times 1.18\) and the expected wheat yield is \(\theta \times 0 + (1 - \theta) \times 1.60\). Barley is favored if \(\theta\) is low and droughts are more likely, wheat is favored if \(\theta\) high, and the value of \(\theta\) at which expected barley and wheat yields are the same is known as the point of indifference (Figure \ref{fig:yields}). If a farmer is unsure whether \(\theta\) is above or below the point of indifference, they can plant a mix of crops that maximizes expected yields given their uncertainty in \(\theta\) (Luce and Raiffa 1989). Thus, the exact risk of a drought occurring in any given year and farmers' perceptions of that risk both have major implications for the decision-making process.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/yields-1.pdf}
\caption{\label{fig:yields}Expected wheat and barley yields under increasing drought risk. The point of indifference is highlighted, at which point planting either wheat or barley results in the same expected yield.}
\end{figure}

\hypertarget{sec:2}{%
\section{Early to mid-Holocene drought risks}\label{sec:2}}

In order to estimate drought risks in the past, present-day weather observations alone are insufficient. Precipitation varies not only from year-to-year, but also on centennial to millennial-scale time scales that are unresolved in the contemporary observational record. Climate dynamics are non-linear, non-stationary, and non-ergodic, which means sudden, unpredictable variability is the norm rather than the exception.

Estimates of past climate variability derived from paleoclimate simulations provides a richer representation of not only the first order statistics of the climate system (e.g.~the mean and variance of precipitation) but also the higher order patterns such as the serial persistence of wet and dry years. These estimates present a more realistic picture of the inherent year-to-year uncertainty in the climate system and presents a more realistic challenge to simple risk-managing strategies that assume climatic risks are fixed.

\hypertarget{paleoclimate-simulation}{%
\subsection{Paleoclimate simulation}\label{paleoclimate-simulation}}

Estimates of changing Holocene precipitation variability were derived from the TraCE-21ka paleoclimate simulation (He 2011). TraCE-21ka is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model to recreate the transient response of the global climate system to changes in the Earth's orbit and greenhouse gas concentrations from the Last Glacial Maximum to the present. The simulation generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of external climate drivers (e.g.~Earth's orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six-hourly timescale, and model outputs are archived at a monthly resolution. For this analysis, monthly TraCE-21ka precipitation outputs were extracted from the 3.75° grid cell covering Central Anatolia. This location was selected to capture climate variability typical for major Neolithic settlements in the region, such as Çatalhöyük and Aşıklı Höyük, and for the eastern Mediterranean more broadly.

\hypertarget{estimating-drought-risks}{%
\subsection{Estimating drought risks}\label{estimating-drought-risks}}

Using the climate model output, I divided each model year into dry years and normal years. A dry year was any year where less than 300mm of rain fell during the wet season (October-May), the threshold below which wheat crops will generally fail (Wilkinson 1997) and a normal year was defined as any year above this threshold. Given the modeled patterns of normal and dry years, the ``objective'' climatic drought risk \(\hat\theta\) for any particular year was defined as the proportion of the previous 50 years that were dry years:
\begin{equation}
    \hat\theta = \frac{\sum_{n=t-1}^{t-50} P_n < 300}{50},
\end{equation}
where \(P_n\) is the growing season precipitation accumulation in millimeters for year \(n\).

The simulated risks of crop failure due to drought ranged between 10\% and 46\% during the period from 9.5ka to 5.5ka, with a median risk of 24\% (Figure \ref{fig:risk}). On average, a Neolithic farmer in Central Anatolia could expect their wheat crops to fail two or three times a decade, punctuated by even drier periods in which wheat crops could be expected to fail roughly every other year. The simulation also reveals a long-term trend of decreasing drought risks, in particular with higher drought risk in the early Holocene giving way to lower drought risk in the middle Holocene.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/risk-1.pdf}
\caption{\label{fig:risk}Annual risk of wheat crop failure due to drought aggregated by fifty-year period. The dashed line indicates the level of risk beyond which one would plant barley over wheat to maximize subjective expected yields (after Figure 1, given the payoffs in Table 1).}
\end{figure}

The TraCE-21ka simulation confirms that drought risk in the eastern Mediterranean was non-stationary and, in fact, quite volatile during periods of climatic disruption in the early Holocene. This volatility would have had severe consequences for early farming communities whose risk-managing practices depended so heavily on accurately perceived local climatic risks.

\hypertarget{sec:3}{%
\section{Modeling risk perception}\label{sec:3}}

In order to properly manage drought risk, a farmer must first be able to perceive that risk. Yet, a farmer's perception of risk reflects more than just the objective, empirical risk observable in the world around them (Tucker 2007; Tucker et al. 2013; Findlater, Satterfield, and Kandlikar 2019). Individual risk perception is inherently subjective, influenced by a person's past experience of dry and wet years as filtered through memory, and can reflect varying levels of uncertainty. Likewise, the distribution of individuals' perceived risks within a population influences the collective perception of drought risks and the potential aggregate societal-level response to those risks (Moussaïd 2013; Amaral and Javarone 2020). But how best to model risk perception at the individual level?

\hypertarget{prior-beliefs-and-bayesian-agents}{%
\subsection{Prior beliefs and Bayesian agents}\label{prior-beliefs-and-bayesian-agents}}

The human brain does not record every bit of perceived information in memory, rather it stores a ``compact encoding'' of that information which it uses for future decision-making (Gallistel et al. 2014). Integrating the tools of agent-based modeling and Bayesian probability provide an elegant means of representing this process. A Bayesian agent is one whose subjective beliefs can be represented as a probability distribution over possible states of nature (Cushman and Macindoe 2009; Pope and Gimblett 2015). This approach has a clear computational efficiency for both modelers and decision-makers. Even if real-world decision makers are not Bayesian in a literal, mathematical, sense, the basic algorithmic problems faced by the brain and the solutions them which it has evolved reflect the same constraints on information processing in the minds of decision-makers (Bonawitz et al. 2011; Sanborn and Chater 2016; Khalvati et al. 2019).

We can examine the dynamics of risk perception using a population of Bayesian agents that develop beliefs about the risk of drought through their personal experience of the weather. These perceptions impact their decision-making by altering their subjective expected crop yields from planting different crop mixes. Here, occurrence of a drought in any given year is treated as a draw from Bernoulli distribution (i.e.~a coin flip) with parameter \(\theta\) representing the drought probability. The beta distribution is a natural choice for representing knowledge about probabilities because it is constrained to fall between 0 and 1. Hence, an individual agent's prior belief about the plausible value of \(\theta\) can be represented as
\begin{equation}
  \theta \sim \textrm{Beta}(\alpha, \,\beta),
\end{equation}
a beta distribution with the parameters \(\alpha\) and \(\beta\) corresponding to the number of dry and wet years previously experienced by that agent. Varying these two parameters thus allows for the representation of a variety of different personal experiences of drought risk. For example, if an agent recalls having lived through 5 dry years and 25 normal years, their prior belief about the chance of a drought occurring in the following year would be represented as a Beta(5, 25) distribution with mean value 0.2 equivalent to the empirical drought risk (\(\hat\theta\)) for that period.

Using probabilities to represent agents' beliefs also allows for estimation of the uncertainty in those beliefs (Figure \ref{fig:prior-prob}). A simple Bayesian agent becomes more certain in their beliefs with age. For example, an agent who experienced 5 wet and 10 normal years and one who experienced 25 wet and 50 normal years would both agree that, on average, droughts occur 50\% of the time. Yet the latter agent would be much more certain in this belief because it is drawn from a larger range of experience (i.e.~a larger sample size). We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of this prior belief. How, then, should an agent update its beliefs in light of new experience?

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/prior-prob-1.pdf}
\caption{\label{fig:prior-prob}Development of an individual's perceived drought risk with time, assuming a fixed drought risk of 0.5. Beliefs are represented as beta distributions, and the increased certainty with age reflects the varying effective sample size of the beta prior.}
\end{figure}

\hypertarget{bayesian-updating-and-the-weight-of-past-experience}{%
\subsection{Bayesian updating and the weight of past experience}\label{bayesian-updating-and-the-weight-of-past-experience}}

As a Bayesian agent must be able to update its beliefs as it moves through time and observes each successive year's weather. It does so by comparing the information in this year's observation with the cumulative weight of their past experience. The agent combines its prior beliefs about drought risk with the likelihood of having observed a drought in the current year in order to generate a posterior distribution representing its updated beliefs about the world.

Crucially, the strength of one's prior beliefs determines how much weight is given to new information (Figure \ref{fig:posterior}). For example, assume two agents -- one aged 15 and the other aged 50 -- who have only ever experienced a normal climate where droughts happen on average two out of every ten years. The mean value of \(\theta\) for both agents would thus be 20\%, but the degree of certainty varies because the older agent bases its inference on many more years of experience than the younger agent. Now, assume the climate suddenly changes such that the drought risk is doubled to 40\% for the next 25 years, not an uncommon occurrence in the simulation of early to mid-Holocene climate. Because the prior beliefs of the agents were so different, their subjective beliefs after the drought are also different even though both experienced the same climate.

For young agents with weak priors, the information of each new year can thus strongly influence their beliefs. But older agents -- having more experience and stronger priors -- will be less likely to update their beliefs when comparing the information from a single year's weather with previous decades' worth of accumulated experience. Although neither agent may perceive the ``true'' climatic drought risk exactly, they nevertheless reflect perfectly rational beliefs about the world. Both agents have rational beliefs and differ only in their prior subjective beliefs. Their relative conservatism or flexibility are not biases, but rather varying perspectives on an inherently uncertain world.

As was the case in the game of fictitious play discussed in Section \ref{sec:1}, a degree of conservatism is warranted when the environment is stable and risks do not vary. But, when the environment is volatile, the ability to change one's mind is crucial (Gallistel et al. 2014). Being too inflexible in one's priors can lead to decisions that are too optimistic when conditions really have changed for the worse. Conversely, inflexible priors can fail to take advantage of good conditions.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/posterior-1.pdf}
\caption{\label{fig:posterior}Change in perceived drought probability in an older (50) and younger (15) agent before and after a 25-year dry period.}
\end{figure}

\hypertarget{sec:4}{%
\section{Risk management and the dynamics of risk perception}\label{sec:4}}
